Clinical advancement forecasting \thanks{\textit{\underline{Citation}}: \textbf{TBD}}

### 2024-04-13 paper/paper.tex
Second draft complete

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{fancyhdr}       % header
\usepackage{graphicx}       % graphics
\usepackage{enumitem}       % lists
\usepackage[table, x11names, svgnames]{xcolor}
\usepackage{subfig}
\usepackage{float}
\usepackage{caption}
\usepackage[section]{placeins}
\usepackage[symbol]{footmisc}
\usepackage{setspace}
\usepackage{overpic}

@@

%% Body
\setlength{\parindent}{20pt} % Set paragraph indentation
\setlength{\fboxsep}{1pt} % Set top and bottom padding around a \colorbox

@@

\begin{abstract}
Choosing which drug targets to pursue for a given disease is one of the most impactful decisions made in the global development of new medicines. This study examines the extent to which the outcomes of clinical trials can be predicted based on a small set of longitudinal (temporally labeled) evidence and properties of drug targets and diseases. We demonstrate a novel statistical learning framework for identifying the top \topRankingPct\% of target-disease pairs that are 4-6x more likely to advance beyond phase 2 trials. This framework is 1.5-2x more effective than an Open Targets composite score based on the same set of evidence. It is also 2x more effective than a common measure for genetic support that has been observed previously, as well as in this study, to confer a 2x higher likelihood of success. Utilizing a subset of our biomedical evidence base, non-negative linear models resulting from this framework can produce simple weighting schemes across various types of human, animal, and cell model genomic, transcriptomic, proteomic, and clinical evidence to identify previously undeveloped target-disease pairs poised for clinical success. In this study we further explore: i) how longitudinal treatment of evidence relates to leakage and reverse causality in biomedical research and how temporalized evidence can mitigate common forms of potential biases and inflation ii) the relative impact of different type of features on our predictions; and iii) an analysis of the space of currently undeveloped, tractable targets predicted with these methods to have the highest likelihood of clinical success. To ease reproduction and deployment, no data is used outside of Open Targets and the described methods require no expert knowledge, and can support expansion of lines of evidence to further improve performance.
% Note: OT multipliers come from Sensitivity / Relative risk ratio distribution across configurations / mean@10 and mean@20
\end{abstract}

@@

It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{PMID:24833294,PMID:35804044,Razuvayevskaya2023.02.07.23285407,PMID:37803084,PMID:36963162}, typically by a factor of at least 2x \cite{Minikel2023.06.23.23291765,Nelson2015-eg,King2019-rc,PMID:30652614,Sadler2023-xd}. This information has been used to devise target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}. It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}. In contrast, we aim to establish a framework for identifying targets-disease (TD) pairs with the very highest possible likelihood of success in order to establish an evidence-based method to optimize research and development pipelines. This is accomplished in this study by integrating a subset of our evidence base from Open Targets \cite{Koscielny2017-rr} spanning human clinical, genomic, transcriptomic and proteomic data as well as cell and animal model evidence, pathway information, and basic literature metrics, in a simple statistical modeling framework.

@@

A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine. This is vital because any method that optimizes for likely \textit{future} clinical success based on \textit{historical} clinical outcomes may easily be biased by the non-random nature with which evidence can be absent. By utilizing "temporalized" evidence -- i.e. evidence for which timing of its emergence can be determined -- and outcomes when training our methods before ultimately applying them to present-day evidence with no restrictions on timing, we reduce the risk of information leaks and inflation of performance estimates. We discuss motivations, prior research and our own analysis on how important this problem is for each source in Section~\ref{sec:results_inflation}.

@@

Precedent studies and methods discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with observed clinical trial success, and 2) clinical trial outcome prediction models. Both are measured against the same outcomes and an important distinction between them lies in how prioritization methods are \textbf{not} directly optimized for those outcomes while outcome prediction methods are. In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial. We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS. Finally, we examine how many present-day target-disease pairs are likely to see success rates matching or exceeding those calibrated benchmarks, that are also undeveloped (i.e. have never been in clinical trials) and have a tractable target.

@@

Information for each of the features in Table~\ref{tab:features} that are denoted as \colorbox{Gainsboro}{temporal} is only used when it was published in a year \textbf{prior} to the first phase 2 trial for a TD pair. Genetic evidence is a notable exception to this rule, and motivations for that exception are detailed in Section~\ref{sec:results_inflation}. A training dataset is then formed by including only TD pairs where the first phase 2 year is between 1990 and 2015. The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a \underline{2 year} offset from the present year (2024) to allow enough time for some trials to complete. While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}. This means a substantial fraction of recent outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation. The distribution of these outcomes, the number of associated targets/diseases and a variety of other statistics on this dataset are presented in Supplementary~Figure~\ref{fig:dataset_statistics}.

@@

The combination of target, disease and target-disease features creates sparsity patterns that are important to understand before interpreting models built from it. Figure~\ref{fig:feature_presence} demonstrates this sparsity by showing that of all the TD pairs entering phase 2 trials for the first time between 2016 and 2022 in our evaluation dataset (N=\evaluationDatasetPairCount), less than 2\% of them ever have evidence directly linking them other than literature co-mentions, which exist for 21\% of those pairs. These TD pairs do, however, very frequently have prior clinical evidence for their associated targets and diseases. Specifically, 8,425 (94\%) have a target and 6,855 (76\%) have a disease that had already been in phase 2 or later trials previously. This is consistent with herding effects observed in recent drug development pipelines \cite{PMID:37117303} over the same time period (2016-2022) and underscores the prognostic value such information may have as it is becoming more and more common and clearly confers lower clinical risk for new drug programs. We also observe that, on top of the clear theoretical, causal relationship between prior target and disease human clinical validation and the likely success of programs for new combinations of such targets and diseases, these features have strong, univariate predictive effects in the evaluation dataset of our study. This is illustrated in Supplementary~Figure~\ref{fig:relative_risk_static_features}, which shows the relative risk of advancement for these features capturing the highest clinical stage previously reached by a target or disease.

Taken together, the paucity of TD pair evidence and the abundance of prior target or disease clinical validation should be considered carefully when interpreting predictive performance. TD pairs predicted to be highly likely to advance clinically despite a lack of target-disease-specific evidence are entirely plausible, but the value of such a prediction is dependent upon the application. We choose to minimize the influence of these cases in our study by focusing on rankings within therapeutic areas that do not extend beyond the number of TD pairs with direct evidence of some kind, as discussed in Section~\ref{sec:methods}. This choice is also reflected in our primary performance metrics, as discussed in Section~\ref{sec:metrics}.

@@

\begin{itemize}
  \item \textbf{RDG}\footnote[2]{\label{ours}These models are constructed in this study rather than being defined by an external source.}: Constrained, L2-regularized linear regressor (a.k.a. "Ridge regressor") fit with all core features
  \item \textbf{RDG-T}\textsuperscript{\ref{ours}}: \textbf{RDG} fit with all features instead of only core features, where the only difference is the inclusion of time since phase 2 transition for a TD pair
  \item \textbf{RDG-X}\textsuperscript{\ref{ours}}: \textbf{RDG} fit \textit{without} human clinical and genetic evidence features
  \item \textbf{GBM}\textsuperscript{\ref{ours}}: Constrained, gradient-boosted machine with fit with all core features
  \item \textbf{GBM-T}\textsuperscript{\ref{ours}}: \textbf{GBM} fit with all features instead of only core features
  \item \textbf{OTS}: Open Targets composite score
\end{itemize}

@@

One key objective of this study is to determine if any model, e.g. RDG, can sort TD pairs with genetic support such that at least some portion of that sorted list has a likelihood of advancement that consistently exceeds what is expected from any one source of genetic support. We find that this goal is met and surpassed by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support and have an equivalent or greater expected rate of advancement. This does not appear to be the case with the OMIM baseline, however the lack of examples in our evaluation dataset with OMIM support makes any determination difficult. See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model. This latter, supplementary figure further emphasizes our focus on prioritizing opportunities beyond those with genetic support and provides examples of TD pairs with multiple lines of evidence.

@@

@@

\begin{enumerate}
  \item The RDG-T model achieves far higher ROC and AP scores through the use of the time since transition feature, which indicates the number of years a TD pair has been underdevelopment after having reached phase 2.
  \item The RDG model, however, matches or exceeds RDG-T in performance among top TD pairs
  \item The RDG-X model, using no human clinical or genetic evidence linked to a disease, outperforms the Open Targets composite score and nearly matches the performance of the RDG model beyond top rankings
  \item Linear models outperform gradient-boosting models by nearly all measures
  \item Constrained linear models outperform unconstrained linear models by nearly all measures
\end{enumerate}

@@

\subsection{Effects}
\label{sec:effects}

The coefficients learned by the RDG model, and the average effects they have across the evaluation dataset, are shown in Figure~\ref{fig:effect_sizes}. This model most highly prioritizes genetic signals that have the greatest coverage, i.e. associations from GWAS studies through the \colorbox{Gainsboro}{ot\_genetics\_portal} feature and associations from any curated clinical genetics source, i.e. EVA, Orphanet, UniProt, Genomics England, ClinGen and gene2phenotype, via the \colorbox{Gainsboro}{curated} \vspace*{0mm} feature.  Notably, literature and target/disease specific clinical features also have substantial effects, followed by indicators of animal evidence and target genetic constraint / expression specificity. Any features not shown were deflated to have no effect, which is possible in this model due to the non-negativity constraint. One such feature worth emphasizing is transcriptomic evidence from Expression Atlas. We found this somewhat surprising, but it is supported by arguments against transcript over/under expression as an indicator of genes that influence disease rather than the other way around \cite{PMID:34561431}.

@@

@@

@@

While it is not possible to address all of these issues, we emphasize that there is a clear pattern across the examples in the list above in that they require \textbf{past} clinical successes and/or failures to arise in the first place. This suggests that accounting for when evidence first emerged would limit the extent of these problems. We do so in this study based solely on publication dates associated with any one piece of information linking target-disease pairs. This also offers a novel opportunity to attempt to quantify what kind of evidence suffers most from these biases. Figure~\ref{fig:evidence_inflation} presents results for this based on a relative risk statistic defined as:

@@

We have demonstrated that simple machine learning methods applied to longitudinal biomedical evidence from many sources can be used to predict clinical outcomes for combinations of drug targets and diseases, without knowledge of molecular properties or trial design details. We have also shown that these methods are more precise in the extremes of their predictions than composite, heuristic scores like those from Open Targets. They also outperform such baselines by more comprehensive, traditional measures of classifier performance; however, we find this less compelling and easier to accomplish than improving performance among the upper tail of the opportunities implied by the very highest predictions. This framework would also support the addition of new lines of evidence over time well as it is designed to automatically determine the relevance of any new information without intervention. Lastly, we find that the space of present-day, undeveloped targets within a disease context that both exceed baseline levels of tractability and have a high predicted likelihood of clinical advancement is substantial. It is likely to grow quickly as well since the breadth of much of the underlying evidence is expanding rapidly \cite{PMID:33214558,PMID:36634672,PMID:31491408}.

@@

An important question, that remains difficult to answer, is how much target prioritization methods like the one explored in this study can improve the efficiency of drug discovery. One 2018 estimate suggests that improving success rates from phase 1 to approval by 1.4x (from 9.6\% to 13.8\%) would reduce the median R\&D cost for a \textbf{single drug} by \$480 million USD \cite{PMID:32125404}. If a 2x increase in success rates from phase 1 to approval could be expected when pursuing only genetically validated TD pairs \cite{Nelson2015-eg}, then this would increase to a \$686 million cost savings per drug. Pushing that even further to 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively. These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_opportunities} there are more than 8,000 such opportunities (i.e. untested, present-day TD pairs) where at least a 3x improvement in success rates could be expected.

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{fancyhdr}       % header
\usepackage{graphicx}       % graphics
\usepackage{enumitem}       % lists
\usepackage{soul}           % underlines
\usepackage[table, x11names, svgnames]{xcolor}
\usepackage{subfig}
\usepackage{float}
\usepackage{caption}
\usepackage[section]{placeins}
\usepackage[symbol]{footmisc}
\usepackage{setspace}
\usepackage{overpic}
\usepackage{titlesec}

@@

%% Body
\setlength{\parindent}{20pt} % Set paragraph indentation
\setlength{\fboxsep}{1pt} % Set top and bottom padding around a \colorbox
\titlespacing\section{0pt}{8pt plus 4pt minus 2pt}{0pt plus 2pt minus 2pt} % Prevent text after section headings from floating to bottom of page

@@

\begin{abstract}
Choosing which drug targets to pursue for a given disease is one of the most impactful decisions made in the global development of new medicines. This study examines the extent to which the outcomes of clinical trials can be predicted based on a small set of longitudinal (temporally labeled) evidence and properties of drug targets and diseases. We demonstrate a novel statistical learning framework for identifying the top \topRankingPct\% of target-disease pairs that are as much as 4-5x more likely to advance beyond phase 2 trials. This framework is 1.5-2x more effective than an Open Targets composite score based on the same set of evidence. It is also 2x more effective than a common measure for genetic support that has been observed previously, as well as in this study, to confer a 2x higher likelihood of success. Utilizing a subset of our biomedical evidence base, non-negative linear models resulting from this framework can produce simple weighting schemes across various types of human, animal, and cell model genomic, transcriptomic, proteomic, and clinical evidence to identify previously undeveloped target-disease pairs poised for clinical success. In this study we further explore: i) how longitudinal treatment of evidence relates to leakage and reverse causality in biomedical research and how temporalized evidence can mitigate common forms of potential biases and inflation ii) the relative impact of different types of features on our predictions; and iii) an analysis of the space of currently undeveloped, tractable targets predicted with these methods to have the highest likelihood of clinical success. To ease reproduction and deployment, no data is used outside of Open Targets and the described methods require no expert knowledge, and can support expansion of lines of evidence to further improve performance.
% Note: OT multipliers come from Sensitivity / Relative risk ratio distribution across configurations / mean@10 and mean@20
\end{abstract}

@@

It has been well established that drugs with human genetic evidence linking their respective targets to indications in clinical trials are more likely to succeed \cite{PMID:24833294,PMID:35804044,Razuvayevskaya2023.02.07.23285407,PMID:37803084,PMID:36963162}, typically by a factor of at least 2x \cite{Minikel2023.06.23.23291765,Nelson2015-eg,King2019-rc,PMID:30652614,Sadler2023-xd}. This information has been used to devise target and target-disease ranking algorithms based primarily on a synthesis of multiple genetic signals alone \cite{PMID:38172303,Koscielny2017-rr,PMID:31253980}. It is also possible to expand the breadth of this genetic support to more targets and diseases based on knowledge graphs, protein interactions and/or disease ontologies \cite{PMID:33262371,Bao2022-bq,Sadler2023-xd,PMID:36087372,PMID:36823319}. \ul{To our knowledge, all such expansion methods identify a larger space of opportunities at the expense of expected success rates}. In contrast, we aim to establish a framework for identifying targets-disease (TD) pairs with the very highest possible likelihood of success in order to establish an evidence-based method to optimize research and development pipelines. This is accomplished in this study by integrating a subset of our evidence base from Open Targets \cite{Koscielny2017-rr} spanning human clinical, genomic, transcriptomic and proteomic data as well as cell and animal model evidence, pathway information, and basic literature metrics, in a simple statistical modeling framework.

@@

\ul{A substantial challenge inherent to building such a system is the need to account for the longitudinal nature of knowledge discovery in biomedicine}. This is vital because any method that optimizes for likely \textit{future} clinical success based on \textit{historical} clinical outcomes may easily be biased by the non-random nature with which evidence can be absent. By utilizing "temporalized" evidence -- i.e. evidence for which timing of its emergence can be determined -- and outcomes when training our methods before ultimately applying them to present-day evidence with no restrictions on timing, we reduce the risk of information leaks and inflation of performance estimates. We discuss motivations, prior research and our own analysis on how important this problem is for each source in Section~\ref{sec:results_inflation}.

@@

Precedent studies and methods discussed so far can largely be categorized as either 1) target and target-disease prioritization methods evaluated based on how well they correlate with observed clinical trial success, and 2) clinical trial outcome prediction models. Both are measured against the same outcomes and an important distinction between them lies in how prioritization methods are \textbf{not} directly optimized for those outcomes while outcome prediction methods are. \ul{In this study, we attempt to bridge these methodologies by predicting clinical trial advancement for target-disease pairs based solely on information that would be present well in advance of any drug program or individual trial}. We then calibrate these predictions to determine what thresholds are necessary to match the observed success rates from benchmarks for genetic support like OMIM \cite{PMID:15608251}, ClinVar \cite{PMID:24234437} and GWAS. Finally, we examine how many present-day target-disease pairs are likely to see success rates matching or exceeding those calibrated benchmarks, that are also undeveloped (i.e. have never been in clinical trials) and have a tractable target.

@@

Information for each of the features in Table~\ref{tab:features} that are denoted as \colorbox{Gainsboro}{temporal} is only used when it was published in a year \textbf{prior} to the first phase 2 trial for a TD pair. Genetic evidence is a notable exception to this rule, and motivations for that exception are detailed in Section~\ref{sec:results_inflation}. A training dataset is then formed by including only TD pairs where the first phase 2 year is between 1990 and 2015. The evaluation dataset then consists of all TD pairs entering phase 2 between 2016 and 2022, with a \ul{2 year} offset from the present year (2024) to allow enough time for some trials to complete. While the average phase 2 trial duration may be as low as 2 years \cite{fdaStepClinical}, other estimates would suggest half of them take longer than 2.9 years \cite{PMID:29394327}. This means a substantial fraction of recent outcomes are censored, that this is an important parameter to test sensitivity to and that time itself is likely to be a crucial covariate in this formulation. The distribution of these outcomes, the number of associated targets/diseases and a variety of other statistics on this dataset are presented in Supplementary~Figure~\ref{fig:dataset_statistics}.

@@

The combination of target, disease and target-disease features creates sparsity patterns that are important to understand before interpreting models built from it. Figure~\ref{fig:feature_presence} demonstrates this sparsity by showing that of all the TD pairs entering phase 2 trials for the first time between 2016 and 2022 in our evaluation dataset (N=\evaluationDatasetPairCount), slightly less than 2\% of them (1.89\%) ever have evidence directly linking them other than literature co-mentions, which exist for 21\% of those pairs. These TD pairs do, however, very frequently have prior clinical evidence for their associated targets and diseases. Specifically, 8,425 (94\%) have a target and 6,855 (76\%) have a disease that had already been in phase 2 or later trials previously. This is consistent with herding effects observed in recent drug development pipelines \cite{PMID:37117303} over the same time period (2016-2022) and underscores the prognostic value such information may have as it is becoming more and more common and clearly confers lower clinical risk for new drug programs. We also observe that, on top of the clear theoretical, causal relationship between prior target and disease human clinical validation and the likely success of programs for new combinations of such targets and diseases, these features have strong, univariate predictive effects in the evaluation dataset of our study. This is illustrated in Supplementary~Figure~\ref{fig:relative_risk_static_features}, which shows the relative risk of advancement for these features capturing the highest clinical stage previously reached by a target or disease.

Taken together, the paucity of TD pair evidence and the abundance of prior target or disease clinical validation should be considered carefully when interpreting predictive performance. TD pairs predicted to be highly likely to advance clinically despite a lack of target-disease-specific evidence are entirely plausible, but the utility of such a prediction may vary depending upon the application. We choose to minimize the influence of these cases in our study by focusing on rankings within therapeutic areas that do not extend beyond the number of TD pairs with direct evidence of some kind, as discussed in Section~\ref{sec:methods}. This choice is also reflected in our primary performance metrics, as discussed in Section~\ref{sec:metrics}.

@@

\begin{itemize}
  \item \textbf{RDG}: Constrained, L2-regularized linear regressor (a.k.a. "Ridge regressor") fit with all core features
  \item \textbf{RDG-T}: \textbf{RDG} fit with all features instead of only core features, where the only difference is the inclusion of time since phase 2 transition for a TD pair
  \item \textbf{RDG-X}: \textbf{RDG} fit \textit{without} human clinical and genetic evidence features
  \item \textbf{GBM}: Constrained, gradient-boosted machine with fit with all core features
  \item \textbf{GBM-T}: \textbf{GBM} fit with all features instead of only core features
\end{itemize}

@@

One key objective of this study is to determine if any model, e.g. RDG, can sort TD pairs with genetic support such that at least some portion of that sorted list has a likelihood of advancement that consistently exceeds what is expected from any one source of genetic support. We find that this goal is met and surpassed by the RDG model, which actually identifies more TD pairs than those that have either EVA or GWAS support and have an equivalent or greater expected rate of advancement. See Section~\ref{sec:results_opportunities} for more on how these benchmarks are employed to contextualize opportunities among undeveloped TD pairs and Supplementary~Figure~\ref{fig:top_evaluation_predictions} for top predictions from the RDG model. This latter, supplementary figure further emphasizes our focus on prioritizing opportunities beyond those with genetic support and provides examples of TD pairs with multiple lines of evidence.

@@

\begin{enumerate}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The RDG-T model achieves far higher ROC and AP scores through the use of the time since transition feature, which indicates the number of years a TD pair has been underdevelopment after having reached phase 2.
  \item The RDG model, however, matches or exceeds RDG-T in performance among top TD pairs
  \item The RDG-X model, using no human clinical or genetic evidence linked to a disease, outperforms the Open Targets composite score and nearly matches the performance of the RDG model beyond top rankings
  \item Linear models outperform gradient-boosting models by nearly all measures
  \item Constrained linear models outperform unconstrained linear models by nearly all measures
\end{enumerate}

@@

@@

\subsection{Feature effects}
\label{sec:effects}

The coefficients learned by the RDG model, and the average effects they have across the evaluation dataset, are shown in Figure~\ref{fig:effect_sizes}. This model most highly prioritizes genetic signals that have the greatest coverage, i.e. associations from GWAS studies through the \colorbox{Gainsboro}{ot\_genetics\_portal} feature and associations from any curated clinical genetics source, i.e. EVA, Orphanet, UniProt, Genomics England, ClinGen and gene2phenotype, via the \colorbox{Gainsboro}{curated} \vspace*{0mm} feature.  Notably, literature and target/disease specific clinical features also have substantial effects, followed by indicators of animal evidence and target genetic constraint or expression specificity. Any features not shown were deflated to have no effect, which is possible in this model due to the non-negativity constraint. One such feature worth emphasizing is transcriptomic evidence from Expression Atlas. We found this somewhat surprising, but it is supported by arguments against transcript over/under expression as an indicator of genes that influence disease rather than the other way around \cite{PMID:34561431}.

@@

Utilizing our predictive models, we systematically identify target-disease pairs that have not yet been explored in clinical trials but exhibit a high likelihood of successful outcomes. This process involved ranking pairs based on their predicted success rates and filtering those with substantial supporting evidence but no prior clinical development attempts.

@@

To address potential biases and inflation in our dataset, our methodology specifically considers the temporal emergence of evidence. By doing so, we ensure our predictions are grounded in data reflective of genuine biomedical insights, rather than artefacts of research focus or historical data accumulation trends.

@@

While it is not possible to address all of these issues, we emphasize that there is a clear pattern across the examples in the list above in that they require \textbf{past} clinical successes and/or failures to arise in the first place. \ul{This suggests that accounting for when evidence first emerged would limit the extent of these problems}. We do so in this study based solely on publication dates associated with any one piece of information linking target-disease pairs. \ul{This also offers a novel opportunity to attempt to quantify what kind of evidence suffers most from these biases}. Figure~\ref{fig:evidence_inflation} presents results for this based on a relative risk statistic defined as:

@@

We have demonstrated that simple machine learning methods applied to longitudinal biomedical evidence from many sources can be used to identify combinations of drug targets and diseases that are 4-5x more likely to advance past phase 2 trials, and that this can be done without knowledge of molecular properties or trial design details. We have also shown that these methods are more precise in the extremes of their predictions than composite, heuristic scores like those from Open Targets. They also outperform such baselines by more comprehensive, traditional measures of classifier performance; however, we find this less compelling and easier to accomplish than improving performance among the upper tail of the opportunities implied by the very highest predictions. This framework would also support the addition of new lines of evidence over time well as it is designed to automatically determine the relevance of any new information without intervention. Lastly, we find that the space of present-day, undeveloped targets within a disease context that both exceed baseline levels of tractability and have a high predicted likelihood of clinical advancement is substantial. It is likely to grow quickly as well since the breadth of much of the underlying evidence is expanding rapidly \cite{PMID:33214558,PMID:36634672,PMID:31491408}.

@@

An important question, that remains difficult to answer, is how much target prioritization methods like the one explored in this study can improve the efficiency of drug discovery. One 2018 estimate suggests that improving success rates from phase 1 to approval by 1.4x (from 9.6\% to 13.8\%) would reduce the median R\&D cost for a \textbf{single drug} by \$480 million USD \cite{PMID:32125404}. If a 2x increase in success rates from phase 1 to approval could be expected when pursuing only genetically validated TD pairs \cite{Nelson2015-eg}, then this would increase to a \$686 million cost savings per drug. Pushing that even further to a 3x or 4x increase by incorporating more lines evidence, as we have done here, would suggest a per-drug cost savings of \$1.03 billion and \$1.37 billion, respectively. These savings would only be realized for drugs tested against targets and diseases with sufficient supporting evidence; however, we show in Section~\ref{sec:results_opportunities} there are more than 8,000 such opportunities (i.e. untested, present-day TD pairs) where at least a 3x improvement in success rates could be expected.
